### 2023-06-22 tex/semantics.tex
[ writing ] expand and reorganise examples

We can now state the data required to implement a traversal assigning
semantics to terms. For open families $\V$ and $\C$, interpreting
variables and terms respectively, we assume that $\V$ is renameable,
that $\V$ is embeddable in $\C$, and that we have an algebra for a
layer of syntax, where bound variables are handled using the Kripke
function space:

@@

>>>>

We can now state the data required to implement a traversal assigning
semantics to terms. For open families $\V$ and $\C$, interpreting
variables and terms respectively, we assume that $\V$ is renameable
(i.e., that $\sdtstile{}\V A \rightarrowtriangle \Box\plr{\sdtstile{}\V A}$ for
all $A$),
that $\V$ is embeddable in $\C$, and that we have an algebra for a
layer of syntax, where bound variables are handled using the Kripke
function space:

@@

\AgdaFunction{semantics} is the fundamental lemma of the framework.
With it proven, I move onto corollaries and specific applications.

\section{Reifying the Kripke function space}

A result I will use throughout the rest of this thesis is \emph{reification}.
When we have an index-preserving mapping from usage-checked variables to
$\V$-environments, we can construct environments of the form
$\Gamma \env\V \Gamma$ (identity environments) for all $\Gamma$.
This lets us write the \AgdaFunction{reify} function, which  simplifies our
obligations in giving a \AgdaRecord{Semantics} by coercing
\AgdaFunction{Kripke} functions into just
\AgdaBound{$\C$}-values in an extended context.

\begin{lemma}[\AgdaFunction{reify}]\label{thm:reify}
  If $\V$ is an open family such that there is a function
  $v : {\sqni} \rightarrowtriangle \V$, we get a function of type
  $\mathrm{Kripke}\,\V\,\C \rightarrowtriangle \mathrm{Scope}\,\C$ for any $\C$.
\end{lemma}
\begin{proof}
  Let $b : \mathrm{Kripke}\,\V\,\C\,\Delta\,\Gamma\,A$.
  That is, $b$ is a Kripke function yielding $\C$-computations
  We want to apply $b$ so as to get a $\C\,\plr{\Gamma, \Delta}\,A$.
  Let $\grP\gamma = \Gamma$ and $\grQ\delta = \Delta$.
  The $\Box^r$ in the type of $b$ allows us to reverse-rename $\Gamma$ to
  $\Gamma, \gr0\delta$.
  Then we give the $\wand$-function an argument in context
  $\gr0\gamma, \Delta$, noting that
  $\plr{\Gamma, \gr0\delta} + \plr{\gr0\gamma, \Delta} = \plr{\Gamma, \Delta}$,
  as we wanted from the result.
  The argument needs type $\gr0\gamma, \Delta \env\V \Delta$.
  We produce this via \cref{thm:env-postren} from an environment
  $\rho : \gr0\gamma, \Delta \env\V \gr0\gamma, \Delta$ created using $v$
  and a renaming which is the complement to that used on $\Box^r$.
\end{proof}

All of the $\V$s used in examples in this paper support identity environments.
However, \citet[p.~27]{AACMM21} give some important examples that do not
support identity environments, and thus cannot use \AgdaFunction{reify}
(\cref{thm:reify}).
The feature that causes the lack of support for identity environments is that
a semantics can make use of the fact that only variables of particular kinds
are bound by the syntax.
In the examples of \citeauthor{AACMM21}, a bidirectionally typed language only
binds variables that synthesise their type, as opposed to those whose type is
checked.
The semantics of type-checking and elaboration rely on variables synthesising
their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
only binds such synthesising variables.
Similar phenomena would appear in, say, a call-by-value language where
variables are values (not computations), or a polarised language where
variables always have a polarity matching their type.

\section{Renaming and substitution}\label{sec:kit-to-sem}

The final completely syntax-generic result I present is simultaneous
substitution.
I derive this as I did in the simply typed case in \cref{sec:gen-sem}:
I first show that a syntactic kit can be turned into a semantics, and then by
instantiating the notion of kit for, in turn, renaming and substitution, the
general semantic traversal gives the result we want.

The notion of \AgdaRecord{Kit} is essentially the same as in the simply typed
case, once we allow for changes to the basic definitions of variables, terms,
and environments (in particular, renamings).

\ExecuteMetaData[\Syntactictex]{Kit}

The first two fields of \AgdaRecord{Semantics} derive directly from fields of
\AgdaRecord{Kit}.
Meanwhile, to handle term constructors, we first \AgdaFunction{reify} to get a
collection of traversed subterms, and then use \AgdaInductiveConstructor{`con}
to assemble these subterms into a similarly shaped syntactic form as we started
with.
The \AgdaField{vr} field is used implicitly in \AgdaFunction{reify}, as it is
used to show that $\V$-identity environments exist.

\ExecuteMetaData[\Syntactictex]{kit-to-sem}

The action of a syntactic traversal on logical rules is basically fixed: we
preserve the logical rule and extend the environment with any newly bound
variables according to \AgdaField{vr}.
Meanwhile, the action on variables is relatively unconstrained: we look up the
variable in the environment to get a $\V$-value, then transform that $\V$-value
into a term using \AgdaField{tm}.

The idea of simultaneous renaming is that variables replace variables, whereas
with simultaneous substitution, terms replace variables.
This translates to environments for renaming containing $\sqni$-values
(variables), and environments for substitution containing $\vdash$-values
(terms).

%To implement renaming and substitution for terms, we now just implement
%syntactic kits for variables and terms, respectively.

\ExecuteMetaData[\Syntactictex]{Ren-Kit}

Notice that \AgdaFunction{ren\textasciicircum$\vdash$}, witnessing the fact
that terms are renameable, is a corollary of \AgdaFunction{Ren-Kit}.

\ExecuteMetaData[\Syntactictex]{Sub-Kit}

### 2023-06-22 tex/syntax.tex
[ writing ] expand and reorganise examples

\todo{Maybe partially merge with the later L/nL section.}
Linear/non-Linear Logic has two sorts of types: linear types $A, B, C$ and
intuitionistic types $X, Y, Z$.
I encode these two sorts in the Agda types
\AgdaDatatype{Ty}\AgdaSpace{}\AgdaInductiveConstructor{lin} and
\AgdaDatatype{Ty}\AgdaSpace{}\AgdaInductiveConstructor{int}, respectively.
Using an indexed type lets me take the total space of \AgdaDatatype{Ty}, named
\AgdaFunction{$\Sigma$Ty}, as the evident dependent pair type.
\AgdaFunction{$\Sigma$Ty} becomes the type of types seen by the framework.
The types themselves are, on the linear side,
a tensor unit \AgdaInductiveConstructor{tI},
a tensor product \AgdaInductiveConstructor{\_t$\otimes$\_}, and
a linear function space \AgdaInductiveConstructor{\_t$\multimap$\_};
on the intuitionistic side,
a unit \AgdaInductiveConstructor{t1},
a product \AgdaInductiveConstructor{\_t$\times$\_}, and
an intuitionistic function space \AgdaInductiveConstructor{\_t$\to$\_};
and adjoint type formers \AgdaInductiveConstructor{tF} and
\AgdaInductiveConstructor{tG}.
I also throw in a linear base type \AgdaInductiveConstructor{$\iota$}.

@@

As in the other examples, I define a data type of rule labels
\AgdaDatatype{`LnL} before defining the type system \AgdaFunction{LnL}.
The system \AgdaFunction{LnL} has types \AgdaFunction{$\Sigma$Ty} and usage
annotations $\{\gr0, \gr1, \gr\omega\}$, as we have used before for linear
systems.
For reference, I give a standard presentation of the rules of L/nL in
\cref{fig:LnL-orig}

We want to be able to embed the L/nL judgement forms
$\Theta; \Gamma \vdash_{\mathcal L} A$ and $\Theta \vdash_{\mathcal C} X$, where
$\Theta$ is made of intuitionistic variables and $\Gamma$ is made of linear
variables.
Following allowed usage patterns, intuitionistic variables will all have usage
annotation $\gr\omega$, while linear variables will each be annotated
either $\gr1$ or $\gr0$.
These assignments of usage annotations are enforced wherever the rules bind new
variables.
By usage subsumption, intuitionistic variables may also end up with annotation
$\gr1$ or $\gr0$, but we see this as a (meaningless) refinement of the
$\gr\omega$ annotation, not causing a soundness problem.
The key invariant is that linear variables never get annotation $\gr\omega$.

The framework already tells us what variables and contexts we can
have --- namely that we can make any assumption in any context.
This means that we can propose ill formed sequents like
$\gr1A \vdash X$.
To make sure that this is not a problem, we want to maintain the invariant that
any \emph{derivation} of a well formed sequent contains only well formed
sequents.
I enforce this locally using the $\Box^{0{+}{*}}$ bunched modality:
Every rule targeting an intuitionistic judgement has its premises wrapped in a
$\Box^{0{+}{*}}$ to restrict the usage annotations to $\gr0$ and $\gr\omega$.
Additionally, rules \TirName{$F$i} and \TirName{$G$e}, which target linear
judgements but have intuitionistic premises, similarly get a $\Box^{0{+}{*}}$.
Note that having the annotation $\gr0$ means that well formed intuitionistic
sequents can contain linear variables annotated $\gr0$, but these variables are
not usable.

The rules are presented in Agda code below, and also using bunched inference
rule notation in \cref{fig:LnL-bunched}

@@

\begin{figure}
  \begin{align*}
    \mathcal J \Coloneqq \; \Theta; \Gamma \vdashL A \; \mid \; \Theta \vdashC X
  \end{align*}
  \begin{mathpar}
    \ebrule{%
      \infer0[$\mathcal L$-var]{\Theta; A \vdashL A}
    } \and
    \ebrule{%
      \infer0[$\mathcal C$-var]{\Theta, X \vdashC X}
    } \and
    \ebrule{%
      \infer0[$I$i]{\Theta; {\cdot} \vdashL I}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL I}
      \hypo{\Theta; \Delta \vdashL C}
      \infer2[$I$e]{\Theta; \Gamma, \Delta \vdashL C}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL A}
      \hypo{\Theta; \Delta \vdashL B}
      \infer2[$\otimes$i]{\Theta; \Gamma, \Delta \vdashL A \otimes B}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL A \otimes B}
      \hypo{\Theta; \Delta, A, B \vdashL C}
      \infer2[$\otimes$e]{\Theta; \Gamma, \Delta \vdashL C}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma, A \vdashL B}
      \infer1[$\multimap$i]{\Theta; \Gamma \vdashL A \multimap B}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL A \multimap B}
      \hypo{\Theta; \Delta \vdashL A}
      \infer2[$\multimap$e]{\Theta; \Gamma, \Delta \vdashL B}
    } \and
    \ebrule{%
      \hypo{\Theta \vdashC X}
      \infer1[$F$i]{\Theta; {\cdot} \vdashL FX}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL FX}
      \hypo{\Theta, X; \Delta \vdashL C}
      \infer2[$F$e]{\Theta; \Gamma, \Delta \vdashL C}
    } \and
    \ebrule{%
      \infer0[$1$i]{\Theta \vdashC 1}
    } \and
    \text{(no \TirName{$1$e})} \and
    \ebrule{%
      \hypo{\Theta \vdashC X}
      \hypo{\Theta \vdashC Y}
      \infer2[$\times$i]{\Theta \vdashC X \times Y}
    } \and
    \ebrule{%
      \hypo{\Theta \vdashC X_0 \times X_1}
      \infer1[$\times$e$_i$]{\Theta \vdashC X_i}
    } \and
    \ebrule{%
      \hypo{\Theta, X \vdashC Y}
      \infer1[$\to$i]{\Theta \vdashC X \to Y}
    } \and
    \ebrule{%
      \hypo{\Theta \vdashC X \to Y}
      \hypo{\Theta \vdashC X}
      \infer2[$\to$e]{\Theta \vdashC Y}
    } \and
    \ebrule{%
      \hypo{\Theta; {\cdot} \vdashL A}
      \infer1[$G$i]{\Theta \vdashC GA}
    } \and
    \ebrule{%
      \hypo{\Theta \vdashC GA}
      \infer1[$G$e]{\Theta; {\cdot} \vdashL A}
    }
  \end{mathpar}
  \caption{Linear/non-Linear Logic in traditional sequent notation}
  \label{fig:LnL-orig}
\end{figure}

Later,\todo{When?} I will give translations between this encoding of L/nL and
$\name$, providing some assurance that the correct logic has been encoded.

>>>>

Linear/non-Linear Logic has two sorts of types: linear types $A, B, C$ and
intuitionistic types $X, Y, Z$.
I encode these two sorts in the Agda types
\AgdaDatatype{Ty}\AgdaSpace{}\AgdaInductiveConstructor{lin} and
\AgdaDatatype{Ty}\AgdaSpace{}\AgdaInductiveConstructor{int}, respectively.
Using an indexed type lets me take the total space of \AgdaDatatype{Ty}, named
\AgdaFunction{$\Sigma$Ty}, as the evident dependent pair type.
\AgdaFunction{$\Sigma$Ty} becomes the type of types seen by the framework.
The types themselves are, on the linear side,
a tensor unit \AgdaInductiveConstructor{tI},
a tensor product \AgdaInductiveConstructor{\_t$\otimes$\_}, and
a linear function space \AgdaInductiveConstructor{\_t$\multimap$\_};
on the intuitionistic side,
a unit \AgdaInductiveConstructor{t1},
a product \AgdaInductiveConstructor{\_t$\times$\_}, and
an intuitionistic function space \AgdaInductiveConstructor{\_t$\to$\_};
and adjoint type formers \AgdaInductiveConstructor{tF} and
\AgdaInductiveConstructor{tG}.
I also include a linear base type \AgdaInductiveConstructor{$\iota$}.

@@

As in the other examples, I define a data type of rule labels
\AgdaDatatype{`LnL} before defining the type system \AgdaFunction{LnL}.
The system \AgdaFunction{LnL} has types \AgdaFunction{$\Sigma$Ty} and usage
annotations $\{\gr0, \gr1, \gr\omega\}$, as I have used before for linear
systems.
For reference, I give a standard presentation of the rules of L/nL in
\cref{fig:LnL-orig}.

\begin{figure}
  \begin{align*}
    A, B, C &\Coloneqq I \mid A \otimes B \mid A \multimap B \mid FX \\
    X, Y, Z &\Coloneqq 1 \mid X \times Y \mid X \to Y \mid GA \\
    \mathcal J &\Coloneqq \; \Theta; \Gamma \vdashL A \; \mid \; \Theta \vdashC X
  \end{align*}
  \begin{mathpar}
    \ebrule{%
      \infer0[$\mathcal L$-var]{\Theta; A \vdashL A}
    } \and
    \ebrule{%
      \infer0[$\mathcal C$-var]{\Theta, X \vdashC X}
    } \and
    \ebrule{%
      \infer0[$I$i]{\Theta; {\cdot} \vdashL I}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL I}
      \hypo{\Theta; \Delta \vdashL C}
      \infer2[$I$e]{\Theta; \Gamma, \Delta \vdashL C}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL A}
      \hypo{\Theta; \Delta \vdashL B}
      \infer2[$\otimes$i]{\Theta; \Gamma, \Delta \vdashL A \otimes B}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL A \otimes B}
      \hypo{\Theta; \Delta, A, B \vdashL C}
      \infer2[$\otimes$e]{\Theta; \Gamma, \Delta \vdashL C}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma, A \vdashL B}
      \infer1[$\multimap$i]{\Theta; \Gamma \vdashL A \multimap B}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL A \multimap B}
      \hypo{\Theta; \Delta \vdashL A}
      \infer2[$\multimap$e]{\Theta; \Gamma, \Delta \vdashL B}
    } \and
    \ebrule{%
      \hypo{\Theta \vdashC X}
      \infer1[$F$i]{\Theta; {\cdot} \vdashL FX}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL FX}
      \hypo{\Theta, X; \Delta \vdashL C}
      \infer2[$F$e]{\Theta; \Gamma, \Delta \vdashL C}
    } \and
    \ebrule{%
      \infer0[$1$i]{\Theta \vdashC 1}
    } \and
    \text{(no \TirName{$1$e})} \and
    \ebrule{%
      \hypo{\Theta \vdashC X}
      \hypo{\Theta \vdashC Y}
      \infer2[$\times$i]{\Theta \vdashC X \times Y}
    } \and
    \ebrule{%
      \hypo{\Theta \vdashC X_0 \times X_1}
      \infer1[$\times$e$_i$]{\Theta \vdashC X_i}
    } \and
    \ebrule{%
      \hypo{\Theta, X \vdashC Y}
      \infer1[$\to$i]{\Theta \vdashC X \to Y}
    } \and
    \ebrule{%
      \hypo{\Theta \vdashC X \to Y}
      \hypo{\Theta \vdashC X}
      \infer2[$\to$e]{\Theta \vdashC Y}
    } \and
    \ebrule{%
      \hypo{\Theta; {\cdot} \vdashL A}
      \infer1[$G$i]{\Theta \vdashC GA}
    } \and
    \ebrule{%
      \hypo{\Theta \vdashC GA}
      \infer1[$G$e]{\Theta; {\cdot} \vdashL A}
    }
  \end{mathpar}
  \caption{Linear/non-Linear Logic in traditional sequent notation}
  \label{fig:LnL-orig}
\end{figure}

We want to be able to embed the L/nL judgement forms
$\Theta; \Gamma \vdashL A$ and $\Theta \vdashC X$, where
$\Theta$ is made of intuitionistic variables and $\Gamma$ is made of linear
variables.
I take the subscript on $\vdashL$ and $\vdashC$ to be a redundant annotation of
whether the conclusion type is \AgdaInductiveConstructor{lin} or
\AgdaInductiveConstructor{int}, so it does not appear in the Agda code, but I
retain it in all non-mechanised presentations.

I will generally ensure that all intuitionistic variables will all have usage
annotation $\gr\omega$, while linear variables will each be annotated
either $\gr1$ or $\gr0$.
These assignments of usage annotations are enforced wherever the rules bind new
variables.
By usage subsumption, intuitionistic variables may also end up with annotation
$\gr1$ or $\gr0$, but we see this as a (meaningless) refinement of the
$\gr\omega$ annotation, not causing a soundness problem.
The key invariant is that linear variables never get annotation $\gr\omega$.

The key difference in form between linear and intuitionistic sequents is that
intuitionistic sequents cannot contain linear variables.
However, syntax descriptions give us no ability to directly state such a
restriction on the kinds of variables.
Therefore, any encoded syntax will allow us to propose ill formed sequents like
$\gr1A \vdashC X$, where a linear variable is to be used for an intuitionistic
conclusion.
To make sure that this is not a problem I use the $\Boxzpt$ bunched modality to
guard many of the rules, making sure that variables with usage annotation $\gr1$
cannot occur in \emph{provable} intuitionistic sequents.
Specifically, every rule targeting an intuitionistic judgement has its premises
wrapped in a $\Boxzpt$.
Additionally, rules \TirName{$F$i} and \TirName{$G$e}, which target linear
judgements but have intuitionistic premises, similarly get a $\Boxzpt$.
Note that having the annotation $\gr0$ means that well formed intuitionistic
sequents can contain linear variables annotated $\gr0$, but these variables are
not usable.

For ease of \emph{producing} terms, we may prefer a ``garbage in; garbage out''
style with a minimal number of uses of $\Boxzpt$, which could be achieved in two
different ways:
\begin{itemize}
  \item
    Ensure that no sequent of the form $\Gamma, \gr1A \vdashC X$ is derivable.
    This leaves boxes only on rules \TirName{$1$-I} and \TirName{$G$-I} --- the
    two logical rules where an intuitionistic conclusion is derived without any
    intuitionistic premises.
  \item
    Ensure that any rule targeting a well formed conclusion has only well formed
    premises.
    This leaves boxes only on rules \TirName{$F$-I} and \TirName{$G$-E} --- the
    two rules where a linear conclusion is derived from intuitionistic premises.
\end{itemize}
It would be interesting to show the equivalence of these two approaches with the
one I have taken in this thesis, but that is left to future work.

The rules are presented in Agda code below, and also using bunched inference
rule notation in \cref{fig:LnL-bunched}.

@@

In \cref{sec:lnl}, I will give translations between the original L/nL system,
this encoding of L/nL, and $\name$, providing some assurance that the correct
logic has been encoded.
